### 2023-03-02 Epidemics_submission/outbreak_response_models_review_v2.tex

Search ((((((Epidemic OR Outbreak OR Emergency OR Reactive OR Crisis))) AND ((Response OR Management OR Control OR Intervention OR Strategies))) AND ((Stochastic OR Transmission OR Computational OR Mathematical OR Mechanistic OR Statistical OR Simulation OR "In silico" OR Dynamic*))) AND model*) AND ((Cholera OR Dengue OR Diphtheria OR Ebola OR "Foot-and-mouth" OR "foot and mouth" OR FMD OR "Hepatitis A" OR "Hepatitis B" OR "Hepatitis E" OR "Haemophilus influenzae type b" OR Hib OR "Human papillomavirus" OR HPV OR Influenza OR "Japanese encephalitis" OR Malaria OR Measles OR "Meningococcal meningitis" OR Mumps OR Pertussis OR "Whooping cough" OR "Pneumococcal disease" OR Poliomyelitis OR Polio OR Rabies OR Rotavirus OR Rubella OR Tetanus OR "Tick-borne encephalitis" OR Tuberculosis OR Typhoid OR Varicella OR Chickenpox OR "Yellow Fever" OR "vaccine-preventable"))

\textbf{Web of Science (Topic search)}:

TOPIC: (Epidemic OR Outbreak OR Emergency OR Reactive OR Crisis) AND TOPIC: (Respon* OR  Manage*  OR  Control  OR  Interven*  OR  Strateg*)) AND TOPIC: (Stochastic OR Transmission OR Computational OR Mathematical OR Mechanistic OR Statistical OR Simulation OR In silico OR Dynamic*) AND TOPIC: (model*) AND TOPIC: (Cholera OR Dengue OR Diphtheria OR Ebola OR "Foot-and-mouth" OR "foot and mouth" OR FMD OR "Hepatitis A" OR "Hepatitis B" OR "Hepatitis E" OR "Haemophilus influenzae type b" OR Hib OR "Human papillomavirus" OR HPV OR Influenza OR "Japanese encephalitis" OR Malaria OR Measles OR "Meningococcal meningitis" OR Mumps OR Pertussis OR "Whooping cough" OR "Pneumococcal disease" OR Poliomyelitis OR Polio OR Rabies OR Rotavirus OR Rubella OR Tetanus OR "Tick-borne encephalitis" OR Tuberculosis OR Typhoid OR Varicella OR Chickenpox OR "Yellow Fever" OR "vaccine-preventable")

\subsection*{Study selection}
We used Endnote version X7.8 and Rayyan web application (\url{https://www.rayyan.ai/}) to combine the results from the three databases and to identify and remove duplicate records. The unique records were exported into Rayyan web application, where they were screened in two stages by three reviewers (JMA, XP, and EBA).

Stage one screening involved excluding studies based on their title and abstract, using the questionnaire that follows. The reviewers only used the information provided in the title and abstract of each study to decide whether it was eligible for inclusion (``Include''), not eligible (``Exclude''), or likely to be included (``Maybe''). The ``Include'' and ``Maybe'' category of studies further went through stage two screening described ahead, which was more stringent.

\subsubsection*{Questionnaire for title and abstract screening (Stage 1)}
\begin{enumerate}
	\item	Is this article written in English?
	\begin{itemize}
		\item 	No: Exclude. Reason: Not English
	\end{itemize}
	\item 	Does the title of this article fit the scope of this review?
	\begin{itemize}
		\item 	No: Exclude. Reason: Title out of scope
	\end{itemize}
	\item	Does the topic of the abstract fit the scope of this review?
	\begin{itemize}
		\item 	No: Exclude. Reason: Topic out of scope
	\end{itemize}
	\item Is this study entirely a review (literature review, systematic review, scoping review, not indicated)? 
	\begin{itemize}
		\item 	Yes: Exclude. Reason: Review
	\end{itemize}
	\item Is any part of this study about an outbreak that happened in the past, was ongoing at the time of the study, or a hypothetical one, including one that could happen in the future? We define an outbreak as a new and sudden rise in the number of cases of a disease in a population, which when left uncontrolled, could lead to large scale geographic spread.
	\begin{itemize}
		\item 	No: Exclude. Reason: Not an outbreak
	\end{itemize}
	\item Is any part of this study about a human infectious disease (Table 1 in \nameref{S1_File}), Ebola or foot-and-mouth disease in livestock?
	\begin{itemize}
		\item 	No: Exclude. Reason: Disease not in scope
	\end{itemize}
	\item Does this study assess the impact – epidemiological or operational - of a real or hypothetical intervention that was mounted or could potentially be mounted in response to an outbreak of the disease in question? Note that we define an assessment as an evaluation of either the absolute or relative impact of an intervention on one of several outcomes including number/proportion of population reached with the intervention (coverage), and a change in size - number of people, duration, spatial - of the outbreak.
	\begin{itemize}
		\item 	No: Exclude. Reason: Not an outbreak intervention assessment
		\item 	Maybe. Reason: Intervention details unclear
	\end{itemize}
	\item Does this study solely use static methods for evaluating the intervention? Static methods include surveys, regression methods, descriptive, and exploratory statistical methods.
	\begin{itemize}
		\item 	No: Exclude. Reason: Not a model
	\end{itemize}
	\item Does this study use any kind of equation, system of equations, or computer simulation to capture the disease’s transmission process or natural history over time?
	\begin{itemize}
		\item 	No: Exclude. Reason: Model not mechanistic
		\item 	Maybe. Reason: Likely a model
	\end{itemize}
\end{enumerate}

Stage one was first piloted with all three reviewers screening the same 50 studies using Rayyan in ``blind mode''. This was so that the reviewers could not see each other's screening decisions (inclusion/exclusion/maybe). After the pilot screening, the results were compared and any inclusion/exclusion conflicts were discussed and resolved. The pilot phase ensured that all the reviewers were using a consistent screening approach. The remaining studies were then screened in duplicate and all conflicting decisions were resolved at the end through discussions among the reviewers.

Stage 2 involved screening the included studies from stage 1, using their full text, in duplicate and with the following questionnaire.

\subsubsection*{Questionnaire for full text screening (Stage 2)}
\begin{enumerate}
	\item Is this article written in English? (This was necessary because some articles could have English abstracts but non-English full text)
	\begin{itemize}
		\item 	If no, exclude. Reason: Not English	
	\end{itemize}
	\item Is this article a report, commentary or any kind of non-quantitative report?
	\begin{itemize}
		\item 	If yes, exclude. Reason: Article type out of scope
	\end{itemize}
	\item Is the full text readily available?
	\begin{itemize}
		\item 	If no, mark as Maybe. Reason: Full text not available	
	\end{itemize}
	\item Is this study entirely a review (literature review, systematic review, scoping review, not indicated)? 
	\begin{itemize}
		\item 	If yes, exclude. Reason: Review
	\end{itemize}
	\item Is the study or a part of it about an outbreak, real or hypothetical?
	\begin{itemize}
		\item 	If no, exclude. Reason: Not an outbreak
	\end{itemize}
	\item Is the disease a human vaccine-preventable disease (Table 1 in \nameref{S1_File}), Ebola, or foot-and-mouth disease? 
	\begin{itemize}
		\item 	If no, exclude. Reason: Not a listed disease
	\end{itemize}
	\item Does this study solely use static methods for evaluating the intervention? Static methods include surveys, regression methods, descriptive, exploratory statistical methods.
	\begin{itemize}
		\item 	If no, exclude. Reason: Static model or method
	\end{itemize}
	\item Does this study use any kind of equation, system of equations, or computer simulation to capture the disease’s transmission process in a dynamic way?
	\begin{itemize}
		\item 	If no, exclude. Reason: Model not mechanistic
		\item 	If unclear, maybe. Reason: Likely a model
	\end{itemize}
	\item Is the model about within-host dynamics?
	\begin{itemize}
		\item 	If yes, exclude. Reason: Within-host model
	\end{itemize}
	\item Does this study attempt to assess the impact – epidemiological or operational - of a real or hypothetical intervention that was mounted or could potentially be mounted in response to an outbreak of the disease in question? We define an assessment as an evaluation of either the absolute or relative impact of an intervention on one of several outcomes including number/proportion of population reached with the intervention (coverage), and a change in size - number of people, duration, spatial - of the outbreak.
	\begin{itemize}
		\item 	If no, exclude. Reason: Not an outbreak intervention assessment
		\item 	If unclear, maybe. Label: Intervention details unclear
	\end{itemize}
\end{enumerate}

\subsection*{Data extraction}
To extract the data from the studies, we used the data extraction questionnaire implemented with the KoboToolbox web application (Section 3 in \nameref{S1_File}). The questionnaire was first piloted with 10 papers among the three reviewers, who worked in duplicate. All conflicts in terms of extracted data were resolved.

Following the pilot phase, the full list of included studies was shared among three reviewers (JMA, XP, and EBA), who extracted the relevant data independently. The extracted data was cross-checked by one reviewer (JMA). When a discrepancy was found in the extracted data, the reviewer referred to the original paper and resolved the conflict.

We extracted data on the type of publication, author affiliation type (academic/government/non-governmental), country/setting studied, whether at least one author affiliation was situated in the country studied, disease studied, the study objectives (retrospective/prospective impact assessment, timing of the modelling practice with respect to the outbreak (retrospective/prospective/real-time), interventions studied, and whether vaccination was the most impactful intervention when compared as a single intervention (yes/no/inconclusive). A full list of the extracted data items are provided (Section 2 in \nameref{S1_File}).

From the model in each paper, we extracted data on the representation of individuals (compartmental vs individual-based), whether spatial structure was represented (yes/no), model structure (deterministic vs stochastic), model parameterization, and validation. We also extracted a predefined list of outcomes (cases averted, final epidemic size, etc) measured with the models. If an outcome was reported that was not on the pre-defined list, we collected it separately as free text for further analysis. We also extracted data on whether the studies included sensitivity analysis or not. This was to understand how uncertainties in the model inputs were dealt with.  We collected data on whether the studies used real world data, that is, excluding simulated data, and whether the data was openly available for download. We also extracted information on whether the code used for the analysis and visualization were made openly available.

\subsection*{Data analysis and synthesis}
To understand how the papers were distributed in terms of author affiliations, we first tallied the unique combinations of author affiliation types. We then grouped the author affiliations into the following two collaboration types and performed all further analyses by stratifying the variable(s) of interest by them:

\begin{enumerate}
	\item \textit{Purely academic:} papers with only academic author affiliations (academics collaborating with other academics), 
	\item \textit{Mixed:} papers with a mixture of academic, government and NGO affiliations. This collaboration type also included papers with only government or only NGO affiliations. 
\end{enumerate}

We first explored the trend in aggregated publications during 1970-2019 by counting the number of publications in each year irrespective of collaboration type.

To explore how the publications changed in time with respect to the collaboration types, we tallied the number of publications and relative proportion (mixed versus purely academic collaboration) of studies per year. We lumped together the publications from 1970 to 2005 due to the small numbers.

To study how the papers were distributed in terms of geographic locations, we split the studies into those that studied actual versus hypothetical locations and used the subset of studies about actual locations to rank the topmost studied locations (country and continent) by frequency.

We also investigated how connected the authors were to the locations studied. Here, we only considered studies on geographic scales up to actual countries and not larger. All studies about locations larger than a country were dropped from this part of the analysis. We tallied, in terms of collaboration types, the number of studies about actual locations (countries) with or without at least one author affiliation in the studied location. If a paper studied more than one location, we treated each location as a separate instance. This increased the denominators accordingly.

To summarize the conclusions drawn about the impact of vaccination as a single intervention, we grouped the studies into three categories, that is, studies that modelled non-vaccination interventions, those that modelled vaccination in combination with other interventions, and those that modelled vaccination as a single intervention for comparison with other interventions. The question was only applicable to the last category of interventions, so we counted the number of studies that found vaccination to be the most impactful or not. Where vaccination was not found to be the most impactful, we briefly summarised the reasons and modelling assumptions.

For the disease studied, model characteristics (structure, dynamics, and spatial heterogeneity), and modelling practices (parametrization, validation, sensitivity analysis, data and code use and availability), we counted the number of studies per collaboration type and reported the results as percentages and fractions of the total. In counting the number of papers per disease studied and outcome measured, if a study had more than one outcome measured or disease, we treated each instance as unique and counted them separately. Hence, in such cases, the denominator of the reported fraction increased from the total studies we reported in our search results.

To summarise the model outcomes, we stratified by the collaboration type and tallied the outcomes used within each group. We reported the top six most used outcomes by both collaboration types and compared the results between the two.

All analyses were perform in R 4.0.5 ~\cite{R2021}. We have provided a database of all the screened papers and their associated extracted data in a csv file, which can be accessed at \url{https://osf.io/dmvst/?view_only=b50a8d3ec21b4a07b7977d0f56e79fc3}. Additionally, all the analysis tables are provided as a supplement (\nameref{S2_File}).

\section*{Results}
\subsection*{Study selection}
We retrieved a total of $12, 986$ bibliographic records from Scopus, PubMed, and Web of Science (Fig \ref{PRISMA_flowchart}). Using Endnote and Rayyan, we identified and removed $7, 974$ duplicates, resulting in $5, 012$ unique records for title, abstract, and full text screening. We exported the unique records into Rayyan for the screening in two stages. Stage 1 involved a title and abstract screening in duplicate by the three reviewers and resulted $4, 211$ excluded studies. The stage 2 screening, which involved screening the full text, led to $548$ excluded studies.

After the full text screening, $253$ studies remained for the data extraction stage of which $228$ on human vaccine-preventable diseases and $25$ studies were on FMD. The FMD studies were only used for comparisons.

\begin{figure}[!h]
	\includegraphics[scale=0.85]{figs/fig1_PRISMA_flowchart}
	\caption{\bf The PRISMA flow chart. Numbers described here include studies for both the human vaccine-preventable diseases and foot-and-mouth disease.}
	\label{PRISMA_flowchart}
\end{figure}

\subsection*{Publications over time}
Overall, a few outbreak response modelling papers were published between 1970-2005 $(3.9\%; 9/228)$, followed by a marked increase in publications until 2019 $(96.1\%; 220/228)$ (Fig \ref{publications_per_year}).

\begin{figure}[!h]	
 \includegraphics[scale=0.65]{figs/fig2_publications_per_year.png}
	\caption{\bf Number of publications per year (1970-2019).}
	\label{publications_per_year}
\end{figure}

We first categorised the papers according to the unique combinations of author affiliation types (Table \ref{studies_per_author_affiliations_type}).

Overall, papers with only authors from academic institution affiliations were the most common (56.1\%; 128/228). Papers with author affiliations from academic and governmental organization affiliations were the second most common (25.0\%; 57/228), and followed by those with academic and NGO affiliations (7.5\%; 17/228). The least common were those with governmental and NGO affiliations (0.9\%; 2/228).

As explained earlier, we ultimately grouped the author affiliation combinations into two collaboration types. There were more purely academic collaborations (56.1\%; 128/228) than mixed collaborations (43.9\%; 100/228).

\begin{table}[!h]
\centering
\caption{\bf Number of studies per unique combination of author affiliations (human diseases). Here, we also show the grouping of author affiliation combinations into purely academic and mixed collaborations. Purely academic collaborations refer to those authored by only authors with academic institution affiliations whereas mixed collaborations include those authored by a mixture of authors with academic institution affiliations, government institutions, and NGOs or only one of the last two.}
\setlength\arrayrulewidth{1pt} 
\begin{tabular}{|l l c|}
\hline
\textbf{Collaboration type} & \textbf{Author affiliation combination} & \textbf{Number of publications} \\ \hline
Purely academic & academic only  & 128 \\ \hline
Mixed & academic + governmental   & 57 \\ \hline
Mixed & academic + NGO                 & 17  \\ \hline
Mixed & academic + governmental + NGO    & 14 \\ \hline
Mixed & governmental only                & 5 \\ \hline
Mixed & NGO only                       & 5  \\ \hline
Mixed & governmental + NGO               & 2 \\ \hline  \rowcolor{gray!20}
\textbf{Total} &  & \textbf{228} \\  \hline 
\end{tabular}
\label{studies_per_author_affiliations_type}
\end{table}

To investigate the changes in collaboration types over time, we calculated the number and proportion of studies per collaboration type per year (Fig \ref{studies_per_collab_type}). In the past seven years (2013-2019), there was an absolute increase in the number of papers by both collaboration types. However, in the same period, there was no increase in the relative proportion of publications per year of mixed collaborations (\nameref{S1_Fig}).

\begin{figure}[!h]
\includegraphics[scale=0.65]{figs/fig3_total_collabs_per_year_plot.png}
	\caption{\bf Total studies by collaboration type. The period from 1970-2005 has been lumped up due to the lack of publications. Academic collaborations refer to those authored by only authors with academic institution affiliations whereas mixed collaborations include those authored by a mixture of authors with academic institution affiliations, government institutions, and NGOs. Mixed collaborations are represented by the turquoise bars and purely academic collaborations by the brown bars.}
	\label{studies_per_collab_type}
\end{figure}

\subsection*{Locations studied}
Overall, most of the papers were about actual locations $(78.6\%; 195/248)$. Among these, some studied a geographic location spanning more than one country but not classifiable as a continent $(9.7\%; 19/195)$. Among those 19 studies, West Africa was studied the most $(47.4\%; 9/19)$, followed by the whole globe $(36.8\%; 7/19)$, Southeast Asia $(10.5\%; 2/19)$, and the Northern Hemisphere $(5.3\%; 1/19)$.

When aggregated into continents, the Americas were studied the most $(36.4\%; 68/187)$, followed by Asia $(25.1\%;47/187)$, Africa $(24.6\%; 46/187)$, Europe $(13.4\%; 25/187)$, and Oceania $(0.5\%; 1/187)$.

When disaggregated into countries, the United States was the most studied $(21.1\%; 37/176)$, followed by China $(10.8\%; 19/176)$, Canada $(6.2\%; 11/176)$, and Sierra Leone $(5.7\%; 10/176)$.

\subsection*{Connection of authors to the location studied}
We investigated the connection of the authors to the location studied and found that, overall, there were more studies with at least one author connected to the studied location $(75.2\%; 118/157)$ than not $(24.8\%; 39/157)$.

When stratified by collaboration types, mixed collaborations were more likely to have studies with at least one author in the location studied $(83.1\%; 63/77)$ compared with the purely academic collaborations $(67.5\%; 54/80)$.

### 2023-03-02 Epidemics_submission/outbreak_response_models_review_v2.tex

\subsection*{Diseases studied}
Overall, the number of studies per disease was disproportionately distributed (Table \ref{studies_per_disease_and_collaboration_type}). Influenza was the most studied $(57.2\%; 135/236)$, followed by Ebola $(14.4\%; 34/236)$, Dengue $(5.1\%; 12/236)$, and a tie between Cholera $(4.7\%; 11/236)$, and Measles $(4.7\%; 11/236)$.

When the diseases were broken down in terms of collaboration type, there were clear differences in their distribution. Influenza, Dengue, and Measles were more studied by mixed collaborations whereas Ebola and Cholera were more studied by purely academic collaborations.

\begin{table}[!h]
		\setlength\arrayrulewidth{1pt} 
	\centering
	\caption{\bf Number of studies per disease and collaboration type. Percentages are calculated from the row totals. The number of studies making up the percentages are shown in brackets. In counting the number of papers per disease studied, if a study was about more than one disease, we treated each instance as unique and counted them separately, leading to a total greater than the reported number of studies.}
	\label{chap3-table: Number of studies per disease and collaboration type}
	\begin{tabular}{|l c c c|}
		\hline
		\textbf{Disease}         & \textbf{Purely academic} & \textbf{Mixed} & \textbf{Total} \\ \hline
		Influenza                & 52.6\% (71)     & 47.4\% (64) & 135   \\ \hline
		Ebola                    & 70.6\% (24)     & 29.4\% (10) & 35    \\ \hline
		Dengue                   & 50.0\% (6)      & 50.0\% (6)  & 12    \\ \hline
		Cholera                  & 81.8\% (9)      & 18.2\% (2)  & 11    \\ \hline
		Measles                  & 36.4\% (4)      & 63.6\% (7)  & 11    \\ \hline
		Tuberculosis             & 85.7\% (6)      & 14.3\% (1)  & 7     \\ \hline
		Poliomyelitis            & 28.6\% (2)      & 71.4\% (5)  & 7     \\ \hline
		Varicella                & 50.0\% (2)      & 50.0\% (2)  & 4     \\ \hline
		Meningococcal meningitis & 33.3\% (1)      & 66.7\% (2)  & 3     \\ \hline
		Pertussis                & 50.0\% (1)      & 50.0\% (1)  & 2     \\ \hline
		Pneumococcal disease     & 50.0\% (1)      & 50.0\% (1)  & 2     \\ \hline
		Yellow fever             & 50.0\% (1)      & 50.0\% (1)  & 2     \\ \hline
		Hepatitis A             & 100.0\% (1)     & 0.0\% (0)   & 1     \\ \hline
		Rubella                  & 100.0\% (1)     & 0.0\% (0)   & 1     \\ \hline
		Typhoid                  & 100.0\% (1)     & 0.0\% (0)   & 1     \\ \hline
		Hepatitis B              & 0.0\% (0)       & 100.0\% (1) & 1     \\ \hline
		Malaria                  & 0.0\% (0)       & 100.0\% (1) & 1     \\ \hline
		Mumps                    & 0.0\% (0)       & 100.0\% (1) & 1    \\ \hline \rowcolor{gray!20}
		\textbf{Overall}			&	\textbf{55.5\% (131)}   			& \textbf{44.5\% (105)}   & \textbf{100\% (236)} \\
		\hline
	\end{tabular}
	\label{studies_per_disease_and_collaboration_type}
\end{table}

\subsection*{Intervention types and impact of vaccination}
There were more papers about non-vaccination interventions $(48.2\%; 110/228)$, followed by those that modelled vaccination as part of a mix of interventions or do-nothing counterfactuals $(41.7\%; 95/228)$, and those that modelled vaccination as a single intervention for side-by-side comparison with other non-vaccination interventions $(10.1\%; 23/228)$.

The third group of papers that modelled vaccination as a single intervention allowed us to collate conclusions on the sole impact of vaccination. There were approximately the same number of studies in this set of studies belonging to the two collaborations types.

Concerning the conclusions about the sole impact of vaccination as a single intervention, most of the studies found vaccination to be the most impactful single intervention in side-by-side comparison with other non-vaccination interventions $(82.6\%; 19/23)$. Few studies found the vaccination to be less impactful compared to other interventions and for various reasons $(17.4\%; 4/23)$. Influenza was the disease studied in these latter four cases. Isolation, and antivirals were found to be more impactful than vaccination due to reasons including the delay until a strain-specific vaccine is developed to control the disease.

\subsection*{Model structure, spatial heterogeneity, and model dynamics}
There were more compartmental models $(78.5\%; 179/228)$ than agent-based models $(20.2\%; 46/228)$ with no clear difference in preference between the two collaboration types.

Approximately a third of the papers included spatial heterogeneity $(28.9\%; 66/228)$ with no clear difference between the two collaboration types.

Deterministic models were the most common $(62.3\%; 142/228)$ compared to stochastic $(31.6\%; 72/228)$ and hybrid models $(6.1\%; 14/228)$, that is models with both deterministic and stochastic components. Here, mixed collaborations were more likely to use stochastic models $(40.0\%; 40/100)$ than purely academic collaborations $(25.0\%; 32/128)$.

\subsection*{Model parametrization and validation}
The top three most commonly used parametrization methods included: combining literature sources and expert opinion/assumptions, literature sources and fitting to data, and literature sources only (Table \ref{model_parametrization_results}).

The most common parametrization method among mixed collaborations was the combination of literature sources and fitting $(27.3 \%; 35/128)$, whereas the combination of literature and expert opinion was the most common among purely academic collaborations $(28.0\%; 28/100)$.

\begin{table}[!h]
\setlength\arrayrulewidth{1pt} 
\centering
\caption{\bf Model parametrization methods. We define parametrization as the method of determining the parameter values for the model. In the table, ``Literature'' means the model's parameters were obtained from literature sources. ``Expert opinion'' means the values were assumed in consultation with experts of the field. ``Fitted'' means the model’s parameters were obtained through some form of mathematical or statistical fitting to a time series of data. }
\begin{adjustwidth}{-2.10in}{0in}
\begin{tabular}{| p{0.17\textwidth}  p{0.14\textwidth}  p{0.14\textwidth}  p{0.12\textwidth}  p{0.13\textwidth}  p{0.13\textwidth}  p{0.12\textwidth}  p{0.1\textwidth}  p{0.07\textwidth} |} \hline 
\textbf{Collaboration type} & \textbf{Literature and expert opinion} & \textbf{Literature and fitted} & \textbf{Literature} & \textbf{Expert opinion} & \textbf{Literature, expert opinion, and fitted} & \textbf{Fitted} & \textbf{Expert opinion and fitted} & \textbf{Total} \\ \hline
Academic & 27.3\% (35) & 22.7\% (29) & 18.0\% (23) & 13.3\% (17) & 9.4\% (12)  & 7.0\% (9) & 2.3\% (3) & 128 \\ \hline
Mixed & 25.0\% (25) & 28.0\% (28) & 15.0\% (15) & 7.0\% (7) & 19.0\% (19) & 5.0\% (5) & 1.0\% (1) & 100 \\ \hline \rowcolor{gray!20}
Overall & 26.3\% (60) & 25.0\% (57) & 16.7\% (38) & 10.5\% (24) & 13.6\% (31) & 6.1\% (14) & 1.8\% (4) & 228 \\ \hline 
\end{tabular}
\end{adjustwidth}
\label{model_parametrization_results}
\end{table}

Most studies did not perform any form of validation $(63.2\%; 144/228)$. Approximately a third used data to validate their models $(34.6\%; 79/228)$ (Table \ref{model_validation_results}). Mixed collaborations were more likely to validate their model with data or the output of an independent model.

\begin{table}[!h]
	\centering
	\setlength\arrayrulewidth{1pt} 
	\caption{\bf Model validation methods. We define model validation as the method by which the model's performance was measured. ``None'' means no validation was performed, or the model’s output was compared to another output from the same model. ``Data'' means the model was compared to independently observed data. ``Another model’s output'' means the model's output was compared to an independent model's output.}
	\begin{adjustwidth}{-0.00in}{0in}
		\begin{tabular}{| p{0.15\textwidth}  p{0.13\textwidth} p{0.13\textwidth} p{0.13\textwidth} p{0.13\textwidth} p{0.12\textwidth}|}
			\hline
			\textbf{Collaboration type} & \textbf{None} & \textbf{Data} & \textbf{Another model's output} & \textbf{Data and another model} & \textbf{Total} \\ \hline
			Academic & 70.3\% (90) & 28.9\% (37) & 0.8\% (1) & 0.0\% (0) & 128 \\ \hline
			Mixed & 54.0\% (54) & 42.0\% (42) & 3.0\% (3) & 1.0\% (1) & 100 \\ \hline \rowcolor{gray!20}
			Overall & 63.2\% (144) & 34.6\% (79) & 1.8\% (4) & 0.4\% (1) & 228\\ \hline
		\end{tabular}
	\end{adjustwidth}
	\label{model_validation_results}
\end{table}

\subsection*{Data and model simulation code}
More than half of the papers used datasets collected independent of the study for either the model parametrization or validation process $(56.1\%; 128/228)$. These papers were split approximately equally between the two collaboration types.

The use of accessible data was, however, different between the two groups. Purely academic collaborations were more likely to use data that could be accessed in the public domain $(81.0\%; 51/63)$ compared to mixed collaborations $(56.9\%; 37/65)$.

Few papers provided access to the model simulation code $(1.7\%; 4/228)$. Here, if a paper indicated that the authors could be contacted for the code, it was deemed as inaccessible due to the many hurdles with getting the code in time. Some papers reported the computer application/software/program/package used (R, Python, C++, Matlab, and so forth) but we did not collect that information.

\subsection*{Patterns in the foot and mouth disease (FMD) literature}
There were $25$ studies about foot and mouth disease outbreaks. Due to the small number of studies, the absolute differences between the two collaboration types were generally not large enough to be considered (\nameref{S2_File}), hence, we report on the aggregated patterns.

In terms of collaboration types, slightly more than half of the $25$ studies were authored by mixed collaborations $(56.0\%; 14/25)$ than purely academic collaborations $(44.0\%; 11/25)$. Almost all the papers had at least one author in the location studied $(92.0\%; 23/25)$.

Overall, vaccination was modelled as a single intervention for comparison in more than half of the studies $(56.0\%; 14/25)$, followed by vaccination in combination with other interventions $(28.0\%; 7/25)$, and no vaccination $(16.0\%; 4/25)$. Few of the first set of studies found vaccination to be the most impactful as a single intervention $(14.3\%; 2/14)$.

In general, the FMD models had more agent-based $(76.0\%; 19/25)$ and spatially explicit structure $(72.0\%; 18/25)$, largely used stochastic dynamics $(56.0\%; 14/25)$, and performed more sensitivity analyses $(56.0\%; 14/25)$.

In terms of model parametrization, the most common method was the combined use of literature sources and expert opinion $(28.0\%; 7/25)$, followed by fitting to data only $(20.0\%; 5/25)$, and the combined use of literature sources, expert opinion, and fitting to data $(16.0\%; 4/25)$. Here, mixed collaborations were more likely than purely academic collaborations to combine literature sources and expert opinion $(42.9\% 6/14)$. Furthermore, most of the FMD models were not validated $(72.0\%; 18/25)$.

\section*{Discussion}
For the period $1970-2019$, co-authorship by purely academic collaborations formed over $56\%$ of the outbreak response impact modelling studies of human vaccine-preventable diseases. Both collaboration types increased in absolute numbers over the past seven years. Regarding modelling practices between the two collaboration types, mixed collaborations were more likely to: (i) include authors affiliated through their institutions to the location studied, (ii) use more complex modelling practices including stochastic model dynamics, parametrization methods involving a combination of literature sources, expert opinion/assumptions, and fitting to data, and validate their models with data collected independent of the study. Even though this review was about human vaccine-preventable diseases, only 51.8\% of the studies modelled vaccination in some form. Moreover, only 10.1\% of the total studies modelled vaccination for comparison with other non-vaccination strategies in the same analysis. Influenza was disproportionately the most studied disease, followed by Ebola, and Dengue.

There were several differences in models and practices between the human disease literature and foot and mouth disease (FMD). Mixed collaborations dominated the FMD literature, with almost all papers having an author in the location studied. Furthermore, most of the FMD models and modelling practices were more complex involving the use of spatial, stochastic, agent-based models. However, the FMD models were less likely to be validated with real world data.

Collaborations, especially between authors from academic institutions, governmental, and non-governmental organizations responsible for outbreak response, could lead to knowledge transfer and improved decision-making \cite{Muscatello2017}. Moreover, such diversity in modelling collaborations could expand the skill sets needed to solve the complex real world outbreak response needs. However, it is well known that there is a weak link between academic research and decision-making in general, and research findings are often not easily translated into decisions \cite{Choi2005,Deelstra2003,Muscatello2017}. This gap in knowledge translation exists especially between outbreak response researchers and public health decision-makers \cite{Rivers2019,Muscatello2017}. There are examples of the commissioning of modelling by organisations such as the US Centers for Disease Control and Prevention (CDC) and the World Health Organization (WHO) to inform decision-making during past outbreaks of Influenza, Ebola, Zika, and Dengue \cite{Muscatello2017}. However, many decision-makers remain cautious about the uptake of modelling due, in part, to issues about transparency in assumptions, credibility and ease of use of modelling software, and adaptability of the results to other settings and existing policies \cite{Muscatello2017}.

To bridge this knowledge translation gap between modellers and decision-makers, modelling collaborations that deepen the interaction between model developers and decision-making have been proposed \cite{Choi2005,Rivers2019}. This could be achieved through the sharing of expertise so that the academics formulate, parametrize, and validate the models taking into account the expert opinions of the decision-makers so that the findings are more tailored towards implementation. Mixed or interdisciplinary collaborations have been reported to be strongly associated with research impact and translation to decision-making \cite{Deelstra2003}. Hence, outbreak response, which is broadly an operational field, would ideally have modelling groups/collaborations that are comprised of academics and decision-makers to bridge this gap. In this review, we observed an absolute increase in the number of papers published by mixed collaborations (Fig \ref{studies_per_collab_type}). This could suggest an increase in the uptake or recognition of modelling as an outbreak response decision-making tool. Future research to investigate and explain this increase could advance our understanding of the contribution of modelling to outbreak response decision-making.

The inclusion of local experts in outbreak response modelling teams helps with more tailored problem-solving and decision-making and better reception of the decisions made \cite{Abramowitz2015}. In that regard, mathematical modelling collaborations involving local experts or, in this review, locally affiliated co-authors are an important first step towards achieving this. We found that most of the human disease studies had authors with an institutional affiliation in the location studied. We observed the same pattern for FMD albeit at a much higher percentage. However, when the human disease data were disaggregated, the results showed that mixed collaborations had a much higher percentage of studies with at least one author in the studied location compared to purely academic collaborations. Coupling this with the absolute increase in publications by mixed collaborations in the past seven years could suggest an increase in the uptake of modelling as a public health decision-making tool. However, we did not measure this causality. Research that measures the direct impact of mathematical modelling in public health decision-making remains lacking in the literature \cite{Muscatello2017}. Future studies designed to investigate this uptake of models and whether it has had an impact on outbreak response would contribute meaningfully to the field of outbreak response modelling.

Three classes of interventions with respect to vaccination were modelled: non-vaccination (mostly antiviral use and non-pharmaceutical interventions), vaccination mixed with other complementary non-vaccination interventions, and vaccination as a single intervention for side-by-side comparison with other non-vaccination alternatives. The last class of interventions were the only eligible set for our analysis of the conclusions about whether vaccination is always the most impactful intervention when modelling is used as the assessment tool. Even though all the reviewed studies were about human vaccine-preventable diseases, only 10.1\% of the studies modelled vaccination as a single intervention for side-by-side comparison with non-vaccination interventions. This lack of studies assessing the sole impact of vaccination reflects the reality that vaccination is often  implemented alongside other interventions and never alone, making it difficult to assess its sole impact. Nevertheless, mathematical modelling is an ethically viable way to assess the sole impact of vaccination, and hence, an essential tool for decision-making.

Among the studies that modelled vaccination as a single intervention for comparison with other single non-vaccination interventions, vaccination was found to be the most impactful intervention in 82\% of the studies. Four studies out of the 23 that modelled influenza found case isolation \cite{Yasuda2009,Gao2016,Chen2020a} and antiviral prophylaxis and/or therapeutics \cite{Gumel2008} to be more impactful than vaccination assuming those interventions could be implemented immediately, given significantly efficient isolation and large antiviral stockpiles before the outbreak/pandemic. Among these four studies, some common assumptions that influenced how vaccination was implemented included: the late arrival of effective vaccines \cite{Yasuda2009}, low versus high vaccination rates \cite{Gao2016}, the use of partially effective pre-existing vaccines against the current strain reactively \cite{Gumel2009} or pre-emptively \cite{Chen2018}. In all four studies, it was, however, recommended that, vaccination be used in combination with the alternative interventions. Even though antivirals were found to be more impactful in some cases, its prolonged use was discouraged to prevent the development of antiviral resistance \cite{Gumel2009}. On the other hand, there was a high percentage of studies in the FMD literature about the use of vaccination as a single intervention and vaccination was rarely found to be the most impactful intervention.

We found some commonalities and differences in the choice of model structure and dynamics between the two collaboration types in the human disease literature. Mixed collaborations were more likely to use ``complex'' models and practices. We are not making any value judgements with regards to complexity but are only highlighting these differences that might require further investigation. Moreover, it is common to model an outbreak using alternative model choices and assumptions depending on the question being answered \cite{Basu2013}. However, the results must always be interpreted with cognisance of the limitations/assumptions of the model.  The use of more complex models and practices by mixed collaborations is likely due to the fact that often, real world public health policy-related decision-making is operational, requiring finer details in models and approaches to answer the questions posed in a practical way. Our definition of mixed collaborations meant that it may have involved decision-makers. It is, therefore, likely that their need for practical solutions could have influenced higher levels of model details or complexity.

The FMD models were generally more complex than the human disease models. This is not surprising because the nature of FMD spread often requires the inclusion of farm structure, farm connectivity, and demographics to capture the disease's dynamics accurately \cite{Kinsley2018}. Future studies designed to explain the differences in modelling practices of human outbreak response modelling groups might help to explain what we have observed in this review.
